Researchers' perceptions of malaria eradication: findings from a mixed-methods analysis of a large online survey

### 2018-04-07 paper.Rmd
major changes

Quantifying an event's probability and time frame is essential for calculating its expected value. In the case of global malaria eradication, uncertainty regarding feasibility makes it difficult for policymakers and public health practitioners to make fully informed decisions. The opportunity cost of investments in eradication-specific interventions can be high, particularly in contexts with other urgent health priorities. In a large online survey of malaria researchers (n = 844), we query perceptions regarding the likelihood and time frame of eradication. We assess pessimism/optimism (via the proxy of years-to-eradication), broken down by area of expertise, and adjust for selection bias. We also carry out a qualitative analysis of free-form comments provided by 61\% of respondents. Our results show that researchers are largely sceptical of the feasibility of near-term malaria eradication, with most believing that eradication will not be achieved for 50 years or more.
\end{changemargin}
\end{center}

\noindent\fbox{%
    \parbox{\textwidth}{%
        \subsection*{Research Highlights}
        \begin{itemize}
          \item Makes novel use of a "wisdom of crowds" approach
          \item Assesses perceived time frame and likelihood of global malaria eradication among researchers
          \item Identifies challenges pertaining to eradication through qualitative analysis
          \item Highlights the gap between researchers' collective pessimism and institutional optimism
        \end{itemize}
      \subsection*{Keywords}
        Malaria, eradication, elimination, mixed methods, survey, crowdsourcing 
    }%
}
m

@@

Malaria is a parasitic disease transmitted among humans by mosquitoes. The _Plasmodium falciparum_ species accounts for a large majority of the 200 million annual cases as well as the half million annual deaths worldwide [@White2014] [@World2016]. Malaria "elimination", the "interruption of all local transmission of the infection in a country or region" [@Tanner2015] is actively being pursued by dozens of countries around the world, leading to a renewed push for "eradication" (the global elmination of malaria) [@Rabinovich_2017].

@@

Most of the current research on expert opinion regarding the feasibility of malaria eradication focuses on the \emph{how} rather than the \emph{if} and the \emph{when} [@Tanner2015]. The participants in the Malaria Eradication Research Agenda (MalERA) process, in particular, have guided research goals and identified gaps in order for elmination to occur [@Alonso2011] [@Rabinovich_2017]. Though the MalERA authors firmly state that eradication is \emph{not} feasible given the "current tools and state of knowledge", mentions to the time frame are vague ("within the lifetime of young scientists just embarking on their careers" [@Alonso2011] and "in the longer term" [@Rabinovich_2017]) and no mention is made of the perception of the overall probability of achieving eradication.  International programmes, such as the WHO Global Malaria Programme (GMP), have acknowledged the need "to take an official position on how and under what timeline malaria eradication could be achieved" [@WHO2015]. Clarity on timeline and likelihood of eradication could inform policy, and plays a crucial role in the economic analysis of the expected value of malaria elimination initiatives. But achieving clarity is difficult, given the many complex and interacting variables which affect malaria transmission, research funding, and technological development. The question of \emph{how} is complex enough, rendering questions of \emph{if} and \emph{when} even more difficult to answer.

The general objective of eradication serves to inspire, rally funder support, motivate researchers, and focus the efforts of public health practitioners. Proponents of disease eradication point to the success of historical and current campaigns (smallpox and polio, respectively), and highlight the long term benefits in health and wealth to future generations (one often cited figure for smallpox was a cost-benefit ratio of more than 400 to 1 [@Barrett_2006]). However, the opportunity cost of investments in eradication-specific interventions is likely to be high, particularly in contexts of very low resources and many competing health problems. Additionally the impact of malaria on economic indicators is heterogeneous, with singificant impact in areas of Sub-Saharan Africa - where the endemicity is high - and little to no impact in areas of low endemicity [@Cutler_2010] [@Barofsky2015]. Additionally, the “expected value” of an investment in a binary scenario (eradication or not) is a function of the probability of the scenario’s occurrence, the temporal lag of that occurrence, and the uncertainy around both probability of occurrence and temporal lag. This uncertainty makes the "risk" of investing in malaria elimination and eradication high. Therefore, estimating the likelihood and time frame of eradication of malaria is essential for making sound investments in health. But how can we quantify these values given the great deal of uncertainty surrounding eradication?

Many studies have shown value in expert elicitation as a means to reduce uncertainty and inform decision making [@Morgan_2014]. But experts can sometimes be mistaken, a fact implicitly recognized when patients request a "second opinion" from another doctor when facing uncertainty. Additionally, relying on only a small panel of experts for a problem as complex as malaria exposes one to bias, just as an overly concentrated investment portfolio exposes one to more risk. As Sir Francis Galton demonstrated in his famous study in which he showed that the crowds' aggregated estimates of cow's weight formed a quasi-normal distribution centered around the true weight [@Wallis2014] [@Galton1907Voxpopuli], averaging the perceptions of many can be more accurate than taking the opinion of any single expert, since the biases of diverse viewpoints can be complementary and symbiotic.

@@

Participants were asked to quantify (as a percentage) the likelihood of eradication in the next 10, 20, 30, 40, and 50 years. Most participants saw eradication as extremely unlikely in the next 10-30 years, but increasingly likely thereafter. Figure 2 shows the distribution of year-specific likelihood perceptions (panels A-E), as well as an illustration of how both likelihood and uncertainty grow over time (panel F).

Participants were asked to provide the number of years they believe it will take until eradication can be achieved. `r length(which(is.na(df$years[df$responded])))` (`r round(length(which(is.na(df$years[df$responded]))) / nrow(df) * 100, digits = 1)`%) were either blank or uninteligible, whereas `r length(which(!is.na(df$years)))` participants responded. Among respondants, `r length(which(df$years >= 50))` (`r round(length(which(df$years >= 50)) / length(which(!is.na(df$years))) * 100, digits = 1)` %) estimated that it would be 50 or more years until eradication.

@@

In order to de-bias sample selection, we employed Heckman's two-step correction method as well as simple binomial logistic regression model to estimate the likelihood of response as a function of sex and (binned) number of citations. To the extent that the two methods presented nearly identical value, we present here the results of the latter. Having estimated the odds of survey participation, we then used the inverse of the selection model's predictions as weights in a simple linear model to adjust our estimates. We run a separate weighted model to estimate the likelihood of eradication at 10, 20, 30, 40, and 50 years. Figure 3 shows both the aggregated perceived likelihood of eradication over time before and after adjusting for sample selection based on binned number of publications. Our adjusted and unadjusted estimates coincide in the long-term, but diverge sharply in the near-term.

@@

Participants were heterogenous in their perceptions, with all of the 5 pre-determined cut-offs containing the whole range of likelihoods (0-100%), and the free-form time frame responses suggesting that eradication would be achieved in a time-frame ranging from less than a decade to never. Heterogeneity can only be partly explained by area of expertise, as the differences in perception between disciplines were minor and insignificant (see Figure 4).

```{r, fig.height = 3, fig.width = 4, fig.cap = 'Average perceived years to aggregation by discipline'}
areas_df <- areas_df %>% arrange(desc(count))
small_areas <- areas_df$area[1:12]
model_formula <- paste0('years ~ ',
                        paste0(paste0('area_', gsub(' ', '_', small_areas)), collapse = ' + '))
model_formula <- as.formula(model_formula)
fit <- lm(model_formula, data = df)
# Get confidence intervals too
ci <- tidy(confint(fit))
x <- cbind(tidy(fit), ci)
names(x) <- c('term', 'estimate', 'se', 'stat', 'p', 'rownames', 'lwr', 'upr')
x <- x %>%
  dplyr::select(term, estimate,
                p, lwr, upr)
the_intercept <- x$estimate[1]
x <- x[-1,]
for (the_column in c('estimate', 'lwr', 'upr')){
  x[,the_column] <- x[,the_column] + the_intercept
}
x$term <- gsub('area_|TRUE', '', x$term)
x <- x %>%
  dplyr::arrange(estimate)
x$term <- gsub('_', '\n', x$term)
x$term <- Hmisc::capitalize(x$term)
x$term[x$term == 'It'] <- 'Information\ntechnology'

@@

There are several reasons to be skeptical that the combined opinions of malaria researchers is a reliable indicator. First, academic researchers are specialists - their narrow, field-specific view of eradication's feasibility is of arguable reliability, given that they may be unfamiliar with the operational, cultural and "real-world" challenges of malaria eradication. Though crowds have been found to be more "wise" than individuals in many cases, the application of an approach similar to the "wisdom of crowds" is not suitable to all classes of problems [@Mannes_2014]. Crowds can be susceptible to social biases [@Lorenz_2011] (although this survey's anonimity largely protects against this issue), and other biases may come in to play, especially given that our study was of a crowd of "specialists", rather than the population as a whole.

@@

As with Keenan et al's survey of experts regarding the feasibility of eradication of neglected tropical diseases [@Keenan2013], our survey did not delve into whether researchers had clinical or operational experience, nor did it assess opinions of program workers, nor did it explore complexities pertaining to different types of or forms of existence of malaria.

>>>>

Quantifying an event's probability and time frame is essential for calculating its expected value. In the case of global malaria eradication, uncertainty regarding feasibility makes it difficult for policymakers and public health practitioners to forecast the expected value of future benefits. The opportunity cost of investments in eradication-specific interventions can be high, particularly in contexts with other urgent health priorities. In a large online survey of malaria researchers (n = 844), we query perceptions regarding the likelihood and time frame of eradication. We assess pessimism/optimism (via the proxy of years-to-eradication), broken down by area of expertise, and adjust for selection bias. We also carry out a qualitative analysis of free-form comments provided by 61\% of respondents. Our results show that researchers are skeptical of the feasibility of near-term malaria eradication, with most believing that eradication will not be achieved for 50 years or more.
\end{changemargin}
\end{center}

\noindent\fbox{%
    \parbox{\textwidth}{%
        \subsection*{Research Highlights}
        \begin{itemize}
          \item Employs a very large online survey of specialist researchers
          \item Assesses perceived time frame and likelihood of global malaria eradication among researchers
          \item Results suggest that most malaria researchers believe short-term global malaria eradication to be unlikely
          \item Identifies challenges pertaining to eradication through qualitative analysis
        \end{itemize}
      \subsection*{Keywords}
        Malaria, eradication, elimination, mixed methods, survey, crowdsourcing 
    }%
}

@@

Malaria is a parasitic disease transmitted among humans by mosquitoes. The _Plasmodium falciparum_ species accounts for a large majority of the 200 million annual cases as well as the half million annual deaths worldwide [@White2014] [@World2016]. Malaria "elimination", the "interruption of all local transmission of the infection in a country or region" [@Tanner2015] is actively being pursued by dozens of countries around the world, leading to a renewed push for "eradication" ("the worldwide interruption of transmission") [@Hopkins2013] [@Rabinovich_2017] [@TheLancet2011].

@@

Most of the current research on expert opinion regarding the feasibility of malaria eradication focuses on the \emph{how} rather than the \emph{if} and the \emph{when} [@Tanner2015]. The participants in the Malaria Eradication Research Agenda (MalERA) process, in particular, have guided research goals and identified gaps in order for elmination to occur [@Alonso2011] [@Rabinovich_2017]. Though the MalERA authors firmly state that eradication is \emph{not} feasible given the "current tools and state of knowledge", mentions to the time frame are vague ("within the lifetime of young scientists just embarking on their careers" [@Alonso2011] and "in the longer term" [@Rabinovich_2017]) and no mention is made of the perception of the overall probability of achieving eradication.  International programmes, such as the WHO Global Malaria Programme (GMP), have acknowledged the need "to take an official position on how and under what timeline malaria eradication could be achieved" [@WHO2015]. Clarity on timeline and likelihood of eradication could inform policy and forecasting models, and plays a crucial role in the economic analysis of the expected value of malaria elimination initiatives. But achieving clarity is difficult, given the many complex and interacting variables which affect malaria transmission, research funding, and technological development. The question of \emph{how} is complex enough, rendering questions of \emph{if} and \emph{when} even more difficult to answer.

The general objective of eradication serves to inspire, rally funder support, motivate researchers, and focus the efforts of public health practitioners. Proponents of disease eradication point to the success of historical and current campaigns (smallpox and polio, respectively), and highlight the long term benefits in health and wealth to future generation. One often cited figure for smallpox was a cost-benefit ratio of more than 400 to 1 [@Barrett_2006]; a more recent analysis on River Blindness suggests that the costs of elimination (relative to staying in "control mode") in Africa would be far lower even in the short term, thanks to the improvements it would lead to in both treatment times and prevented surveillance costs [@Kim2017].   However, the opportunity cost of investments in eradication-specific interventions is likely to be high, particularly in contexts of very low resources and many competing health problems. Additionally the estimated impact of malaria on economic indicators is heterogeneous, with singificant impact in areas of Sub-Saharan Africa - where the endemicity is high - and little to no impact in areas of low endemicity [@Cutler_2010] [@Barofsky2015]. Additionally, the “expected value” of an investment in a binary scenario (eradication or not) is a function of the probability of the scenario’s occurrence, the temporal lag of that occurrence, and the uncertainy around both probability of occurrence and temporal lag. This uncertainty makes the "risk" of investing in malaria elimination and eradication high. Therefore, estimating the likelihood and time frame of eradication of malaria is essential for making sound investments in health. But how can we prospectively quantify these values given the great deal of uncertainty surrounding eradication?

Time to eradication, a crucial component in any hypothetical cost-benefit analysis that takes into account discounting, is difficult to assess and requires several layers of assumptions in order to generate even a range of possible scenarios; this can be done with a high degree of reliability for diseases closer to eradication such as Lympathic Filariasis [@Kastner2015], but the uncertainties regarding malaria's future treatment protocols and epidemiology are too great to allow for a robust analysis based on cascading assumptions. Many studies have shown value in expert elicitation as a means to reduce uncertainty and inform decision making [@Morgan_2014]. But experts can sometimes be mistaken, a fact implicitly recognized when patients request a "second opinion" from another doctor when facing uncertainty. Additionally, relying on only a small panel of experts for a problem as complex as malaria exposes one to bias, just as an overly concentrated investment portfolio exposes one to more risk. As Sir Francis Galton demonstrated in his famous study in which he showed that the crowds' aggregated estimates of cow's weight formed a quasi-normal distribution centered around the true weight [@Wallis2014] [@Galton1907Voxpopuli], averaging the perceptions of many can be more accurate than taking the opinion of any single expert, since the biases of diverse viewpoints can be complementary and symbiotic.

@@

Most participants saw eradication as extremely unlikely in the next 10-30 years, but increasingly likely thereafter. Figure 2 shows the distribution of year-specific likelihood perceptions (panels A-E), as well as an illustration of how both likelihood and uncertainty grow over time (panel F). At the 40 year mark, the distribution of perceived likelihood of eradication appears "normal", and by 50 years it is slightly shifted to the right (ie, consensus is towards eradication more probable than not).

In regards to responses to perceived years until eradication, `r length(which(is.na(df$years[df$responded])))` (`r round(length(which(is.na(df$years[df$responded]))) / nrow(df) * 100, digits = 1)`%) were either blank or uninteligible, whereas `r length(which(!is.na(df$years)))` participants responded. Among respondants, `r length(which(df$years >= 50))` (`r round(length(which(df$years >= 50)) / length(which(!is.na(df$years))) * 100, digits = 1)` %) estimated that it would be 50 or more years until eradication.

@@

In order to de-bias sample selection, we employed Heckman's two-step correction method as well as simple binomial logistic regression model to estimate the likelihood of response as a function of sex and (binned) number of citations. To the extent that the two methods presented nearly identical value, we present here the results of the latter. Having estimated the odds of survey participation, we then used the inverse of the selection model's predictions as weights in a simple linear model to adjust our estimates. We run a separate weighted model to estimate the likelihood of eradication at 10, 20, 30, 40, and 50 years. Figure 3 shows both the aggregated perceived likelihood of eradication over time before and after adjusting for sample selection based on binned number of publications. Our adjusted and unadjusted estimates coincide in the long-term, but diverge sharply in the near-term; adjustment suggests that had our pool of respondents been more representative, the average perceived likelihood of near-term eradcation would have been much lower.

@@

```{r, fig.height = 3, fig.width = 4, fig.cap = 'Average perceived years to aggregation by discipline', eval = FALSE}

# Participants were heterogenous in their perceptions, with all of the 5 pre-determined cut-offs containing the whole range of likelihoods (0-100%), and the free-form time frame responses suggesting that eradication would be achieved in a time-frame ranging from less than a decade to never. Heterogeneity can only be partly explained by area of expertise, as the differences in perception between disciplines were minor and insignificant (see Figure 4).

areas_df <- areas_df %>% arrange(desc(count))
small_areas <- areas_df$area[1:12]
model_formula <- paste0('years ~ ',
                        paste0(paste0('area_', gsub(' ', '_', small_areas)), collapse = ' + '))
model_formula <- as.formula(model_formula)
fit <- lm(model_formula, data = df)
# Get confidence intervals too
ci <- tidy(confint(fit))
x <- cbind(tidy(fit), ci)
names(x) <- c('term', 'estimate', 'se', 'stat', 'p', 'rownames', 'lwr', 'upr')
x <- x %>%
  dplyr::select(term, estimate,
                p, lwr, upr)
the_intercept <- x$estimate[1]
x <- x[-1,]
for (the_column in c('estimate', 'lwr', 'upr')){
  x[,the_column] <- x[,the_column] + the_intercept
}
x$term <- gsub('area_|TRUE', '', x$term)
x <- x %>%
  dplyr::arrange(estimate)
x$term <- gsub('_', '\n', x$term)
x$term <- Hmisc::capitalize(x$term)
x$term[x$term == 'It'] <- 'Information\ntechnology'

@@

This paper has several limitations. First, academic researchers are specialists - their narrow, field-specific view of eradication's feasibility is of arguable reliability, given that they may be unfamiliar with the operational, cultural and "real-world" challenges of malaria eradication. Though crowds have been found to be more "wise" than individuals in many cases, the application of an approach similar to the "wisdom of crowds" is not suitable to all classes of problems [@Mannes_2014]. Crowds can be susceptible to social biases [@Lorenz_2011] (although this survey's anonimity largely protects against this issue), and other biases may come in to play, especially given that our study was of a crowd of "specialists", rather than the population as a whole.

@@

As with Keenan et al's survey of experts regarding the feasibility of eradication of neglected tropical diseases [@Keenan2013], our survey detected relatively high levels of eradication skepticism and did not delve into whether researchers had clinical or operational experience, nor did it assess opinions of program workers, nor did it explore complexities pertaining to different types of or forms of existence of malaria. Though our sample size was over twice Keenan's, this was largely due to having contacted more authors, as our response rate was only one fourth as large.
